Prosper Loan Data Analysis by Carlos Blanco
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(dplyr)
library(GGally)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
ld <- read.csv('prosperLoanData.csv')

#Select the variables that will be analyzed on a vector
variables_of_interest = c(
'BorrowerRate',
'EstimatedReturn',
'ProsperRating..numeric.',
'EmploymentStatus',
'CreditScoreRangeLower',
'CreditScoreRangeUpper',
'DebtToIncomeRatio',
'IncomeRange',
'LoanOriginalAmount',
'LoanOriginationDate')

#Keep only the data that will be analyzed on the database
ld <- ld[variables_of_interest]
```

This dataset is from prosper and contains data for 113937 loans they made from
2006 to 2014. The main goal is to determine what variables set the interest paid 
by the borrowers (BorrowerRate)

# Univariate Plots Section

To start this analysis is important to fully know the characteristics of the 
data set, next are the variables names, classes and summaries.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
str(ld)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(ld)
```

I observe that I need to transform some variables. First the Prosper rating is 
a number and we will need it as factor, also the loan origination date is in 
datetime format and I proceed to extract years to study the changes over 
time. 

I will combine Credit Score Range Upper and Lower into an Average
because they measure the same variable and rearrange the order of Income Levels
for a better plot visualization.

Also, I observe that a lot of variables have NAs, so in order to examine data
with all of the variables I will remove rows with NAs.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Extract the first 4 characters of LoanOriginationDate because they represent
#the year in the datetime. i.e. 2014-01-22 00:00:00
ld$LoanOriginationDate.year = substr(ld$LoanOriginationDate, 1, 4)

# Combine Credit Score Ranges to obtain the average
ld$CreditScoreAverage <- (ld$CreditScoreRangeLower + ld$CreditScoreRangeUpper)/2

# Since the rating can be considered a categorical value, we will convert it to
#a factor variable to create box plots and help on in visualizations
ld$ProsperRating..factor. <- factor(ld$ProsperRating..numeric.)

#Rearrange Income Range factors to create a incremental order and create
#ordered visualizations
ld$IncomeRange <- factor(
  ld$IncomeRange,
  levels(ld$IncomeRange)[c(8,1,2,4,5,6,3,7)])

#remove rows without data for each variable to detect patterns between them
ld.no_na <- ld[complete.cases(ld),]
```


Next, I looked at the histograms for the BorrowerRate with 3 bins 
sizes to determine the shape of the distribution, I selected bins of 1%, 2,5% 
and 0.5%.


```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram of binsize = 1%
ggplot(data = ld,
       aes(x = BorrowerRate)) + geom_histogram( binwidth = 0.01)

#histogram of binsize = 2.5%
ggplot(data = ld,
       aes(x = BorrowerRate)) + geom_histogram( binwidth = 0.025)

#histogram of binsize = 0.5%
ggplot(data = ld,
       aes(x = BorrowerRate)) + geom_histogram( binwidth = 0.005)
```


In addition, I created a Box Plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld, aes(x = "", y = BorrowerRate)) + 
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0,0.5,0.05))
```


To get a better understanding I plot the mean, median and interquartile 
measures in the 1% bin size histogram.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld,
       aes(x = BorrowerRate)) + 
  geom_histogram( binwidth = 0.01) +
  geom_vline(aes(xintercept = mean(ld$BorrowerRate)),col='red',size=1) +
  geom_vline(aes(xintercept = median(ld$BorrowerRate)),col='blue',size=1) +
  geom_vline(aes(xintercept = quantile((ld$BorrowerRate), 0.25)), linetype= 2) +
  geom_vline(aes(xintercept = quantile((ld$BorrowerRate), 0.75)), linetype= 2)
```


BorrowerRate's summary statistics are:
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(ld.no_na$BorrowerRate)
```


From the previous plot we observe that the distribution is close to a normal one
but has a spike at about 0.32, which moves the mean and the median from the 
center of the curve. This must be further investigated.

Also, next I present the histograms for every variable on the data set in order
to get a better grasp of their distributions.

### Prosper Rating

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = ProsperRating..numeric.)) + 
  geom_histogram() +
  scale_x_continuous(breaks = seq(1,7,1))
```

Observations: It looks close to a normal distribution without tails, because the
ratings only go from 1 to 7.

In addition is benefitial to the analysis the exact
numbers of measures of central tendency. Next we have this measures for each
of the costumbers groups by rating:

Summary Posper Rating == 1
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(ld.no_na$BorrowerRate, ld.no_na$ProsperRating..factor. == 1))
```

Summary Posper Rating == 2
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(ld.no_na$BorrowerRate, ld.no_na$ProsperRating..factor. == 2))
```

Summary Posper Rating == 3
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(ld.no_na$BorrowerRate, ld.no_na$ProsperRating..factor. == 3))
```

Summary Posper Rating == 4
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(ld.no_na$BorrowerRate, ld.no_na$ProsperRating..factor. == 4))
```

Summary Posper Rating == 5
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(ld.no_na$BorrowerRate, ld.no_na$ProsperRating..factor. == 5))
```

Summary Posper Rating == 6
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(ld.no_na$BorrowerRate, ld.no_na$ProsperRating..factor. == 6))
```

Summary Posper Rating == 7
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(ld.no_na$BorrowerRate, ld.no_na$ProsperRating..factor. == 7))
```


### Credit Score Average Histogram

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = CreditScoreAverage)) + 
  geom_histogram(binwidth = 10) +
  scale_x_continuous(breaks = seq(10,850,40)) +
  scale_y_continuous(breaks = seq(0,15000,1000))
```


Summary Credit Score Average
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(ld.no_na$CreditScoreAverage)
```

Observations: Close to normal distribution, moved towards the lower rates. Even 
thus, it is a continuous variable there are bins of size one each 20 points. Also 
the range is 280 (889.5-609.5)


### Loan Origination Date Histogram

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = as.numeric(LoanOriginationDate.year))) + 
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(2005,2014,1))
```

Table of Loan Origination Date's proportions:
```{r echo=FALSE, message=FALSE, warning=FALSE}
prop.table(table(ld.no_na$LoanOriginationDate.year))
```

Observation: in 2013 the company issued the majority of the database loans.


### Debt to Income Ratio Histogram of 99 percentile.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = DebtToIncomeRatio)) + 
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(
    lim = c(0, quantile(ld.no_na$DebtToIncomeRatio, 0.99)),
    breaks = seq(0,0.7,0.05))
```

Observations: Debt to Income Ratio have a normal distribution, however the study
could benefit from the exact measures of central tendency:


Summary Debt to Income Ratio
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(ld.no_na$DebtToIncomeRatio)
```


### Loan Original Amount Histogram

```{r echo=FALSE, message=FALSE, warning=FALSE}
#LoanOriginalAmount is multiplied by 100 to obtain percentages and improve 
#quality of plot
ggplot(data = ld.no_na, 
  aes(x = LoanOriginalAmount / 1000)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(0,35,2))
```

Observations: We can see that there are more loans of 15000 or lower, with peaks
at around 4000, 10000 and 15000. Is there any significate variance through the 
years?


### Estimated return Histogram

```{r echo=FALSE, message=FALSE, warning=FALSE}
#EstimatedReturn is multiplied by 100 to obtain percentages and improve 
#quality of plot
ggplot(data = ld.no_na,
       aes(x = EstimatedReturn * 100)) + 
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(-20,30,2)) + 
  scale_y_continuous(breaks = seq(0,20000,1000))
```

Observations: The bank almost never issues a loan that is expected to bring
negative return, however there are some cases. I thought that the bank will try 
to get standard returns % for each loan, however the distributions seems normal
with the majority of the values ranging from 4% to 16%. 


The exact measures of central tendency are:
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(ld$EstimatedReturn)
```


### Employment Status Count Bar Graph

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = EmploymentStatus)) + geom_bar() +
  scale_y_continuous(breaks = seq(0,65000,5000))
```

Observations: as we suspected not employed people usually don't get a loan, but
neither does retired, self-employed and part-time people.

### Income Range Histogram

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = IncomeRange)) + geom_bar() +
  scale_y_continuous(breaks = seq(0,25000,2500))
```

Observations: the mayority of the loans are made for coustumers who earn between
25000 and 75000, however those who earn more than 75000 also request a good 
amount of loans, the exact numbers are:

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(ld.no_na$IncomeRange)
```


# Univariate Analysis

The dataset is in a wide format, with each row containing the information for a 
single loan, the loan and the borrower characteristics

The main feature of interest is the Borrower Rate, I want to be able to predict
what rate a certain costumer would get.

In order to do this, I will take a closer look at the variables of related to 
the borrowers like credit score, income range, debt to income ratio, the amount 
of the loan and the rating prosper gives to each borrower.

Also, I created a credit average score, because it made sense to 
compare the data to a single credit than an upper and lower score. Also, I 
extracted the year from the loan origination date because I wanted to analyze 
the loans over the years and not a more granular time metric. Finally, I created
a factor variable of the prosper ratings to be able to clearly differentiate the
different borrowers in the plots.

# Bivariate Plots Section

I will use GGpairs as my first plot to get an overview of the relationship
between the variables of study.

In order to improve readability, I will plot BorrowerRate vs with 3 variables at
the time. The first set includes EstimatedReturn, ProsperRating..numeric. and 
ProsperRating..factor.

### GGPairs Plot #1

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create a Dataframe with Borrower Rate and 3 First Variables
variables <- data.frame(
        BorrowerRate = ld.no_na$BorrowerRate,
        EstimatedReturn = ld.no_na$EstimatedReturn,
        ProsperRatingNumeric = ld.no_na$ProsperRating..numeric.,
        ProsperRatingFactor = ld.no_na$ProsperRating..factor.
)

# Plot Dataframe of Borrower Rate and the 3 First Variables
ggpairs(variables, 
        wrap = c(Shape = I("."), outlier.shape = I("."))) + 
        theme(legend.position = "none",
              panel.grid.major = element_blank(),
              axis.ticks = element_blank(),
              axis.text = element_blank(),
              axis.title.y.right = element_text(angle = 45, hjust = 1))
```

GGpairs plot #1 Observations: 
-EstimatedReturn and Prosper Rating Numeric have big correlations with Borrower 
Rate
-BorrowerRate seems to be a normal distribution with a peak in higher levels.
-BorrowerRate seems to decrease as Propsper Rating Factor Increases
-BorrowerRate faceted by Posper Rating Factor seems to create normal 
distribution for each rating
-Prosper Rating Numeric and Factor have normal distributions
-Estimated Return is lower for better Prosper Rating (Beeing 1 the worst and 7
the best)

Next is the plot of EmploymentStatus, IncomeRange and LoanOriginationDate.year 
(numeric).


### GGPairs Plot #2

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create a Dataframe with Borrower Rate and 3 Second Variables
variables <- data.frame(
        BorrowerRate = ld.no_na$BorrowerRate,
        EmploymentStatus = ld.no_na$EmploymentStatus,
        IncomeRange = ld.no_na$IncomeRange,
        LoanOriginationDate.year = as.numeric(ld.no_na$LoanOriginationDate.year)
)

# Plot Dataframe of Borrower Rate and the 3 Second Variables
ggpairs(variables, 
        wrap = c(Shape = I("."), outlier.shape = I("."))) + 
        theme(legend.position = "none",
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())
```

GGpairs plot #2 Observations:
-Employment status does not seems to have a linear relation with BorrowerRate
-The majority of loans are for 1 factor of Employment Status
-As income range increases Borrower Rate summary statistics decreases
-The first Income Rage bin have close to none loan
-LoanOriginationDate.Year have a low correlation with Borrower Rate
-The majority of the loans were made on one year and decreases significantly for
the others.

To conclude the GGpairs plots its the plot of CreditScoreAverage, 
DebtToIncomeRatio and LoanOriginalAmount


### GGPairs Plot #3

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create a Dataframe with Borrower Rate and the 3 third Variables
variables <- data.frame(
        BorrowerRate = ld.no_na$BorrowerRate,
        CreditScoreAverage = ld.no_na$CreditScoreAverage,
        DebtToIncomeRatio = ld.no_na$DebtToIncomeRatio,
        LoanOriginalAmount = ld.no_na$LoanOriginalAmount
)

# Plot Dataframe of Borrower Rate and the 3 third Variables
ggpairs(variables, 
        wrap = c(Shape = I("."), outlier.shape = I("."))) + 
        theme(legend.position = "none",
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())
```

GGpairs plot #3 Observations:
-ECredit Score Average and Loan Original Amount have medium correlations 
(0.4-0.6) with Borrower Rate
-Loan Original Amount histogram have 5 peaks.
-Credit Score average have several peaks, and seems to be a normal distribution
-Debt to Income Ratio has a low correlation with Borrower Rate

To have a better understading of the initial obervations I created plots to
analyze each variable, first for Prosper Rating I created a Scatter Plot vs the
main variable of study Borrower Rate:


####Scatterplot ProsperRating..numeric. vs Borrower Rate

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = ProsperRating..numeric., y = BorrowerRate)) + geom_point()
```

I Observe that there are only round numbers for Prosper Rating, for that reason 
and overplotting I created a boxplot of the same variables to understand the
measures of central tendency.


####Boxplot ProsperRating..factor. vs Borrower Rate

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = ProsperRating..factor., y = BorrowerRate)) + geom_boxplot()
```

The previous plot demostrates how the spread and measures decrease as the rating
increases. This raises the question, how does the rating is separated through 
the Borrower Rate Histogram?


The next plot answers it:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = BorrowerRate)) + 
  geom_histogram(binwidth = 0.01, aes(fill = ProsperRating..factor.)) +
  scale_fill_brewer(type = 'seq', palette = "Spectral")
```

We can confirm that there is a clear separation of Borrower Rate by rating,
also, the peak we saw earlier is conformed only by the costumers with rating 1.
But how is the distribution for each prosper rating? Is it normal if?


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = BorrowerRate)) + 
  geom_histogram(binwidth = 0.005, aes(fill = ProsperRating..factor.)) +
  scale_fill_brewer(type = 'seq', palette = "Spectral") +
  facet_wrap( ~ ld.no_na$ProsperRating..factor.)
```

As we can see in the previous plots, prosper ratings groups but rating 1 have a 
close to normal distributions. 

From the previous plots, we can see that prosper rating is indeed a very
influential variable in the borrower rate. In fact, we can see that the borrower
rate mean, median, interquartile distributions, min and max decrease as the 
rating goes up. 

Also, it is very clear that most of the borrower with rating 1
get approximately 0.31 Rate. Them costumers with rating 2 have a greater spread 
than those with rating 3, and this repeats up to rating 7, where the 
interquartile is very close to the median. 

In addition, when I plotted the 
histograms faceted by rating we can see that all ratings but rating = 1 have a 
close to normal distribution.

Next, I proceed to study the influence of other variables on the Borrower Rate.
For each continuous variable I analyzed the relationship vs Borrower rate and the
variable histogram. After that I answer any particular question that arose from
the previous plots.

### Credit Score Average vs Borrower Rate

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = CreditScoreAverage ,
           y = BorrowerRate)) +
  geom_point(position = 'jitter', alpha = 1/25) + 
  scale_x_continuous(breaks = seq(610, 890, 20)) +
  scale_y_continuous(breaks = seq(0,0.35,0.025))
```


Observations: There seems to be a negative linear relationship, however the 
spread in big and the confidence intervals of a model would be big as well.

If credit score has a negative relationship with Borrower Rate, and also does
the prosper rating, how are they related to each other?


### Credit Score Average vs ProsperRating..numeric.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = CreditScoreAverage, y = ProsperRating..numeric.)) + 
  geom_bin2d() +
  scale_x_continuous(breaks = seq(10,890,20)) +
  scale_y_continuous(breaks = seq(0,7,1))
```


The previous plot does not provide any clear relationship, since for most of the
Credit Scores Averages there are most of the possible Prosper Ratings. 

Next, I analyzed Loan Origination year.


### Loan Origination Date vs Borrower Rate

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = LoanOriginationDate.year,
           y = BorrowerRate)) +
  geom_point(alpha = 1/10, position = 'jitter') +
  scale_x_discrete(breaks = seq(2009,2014,1))
```

Observations: The rates become a little bit lower as the years increases.

The following variable of study is Debt to Income Ratio


### Debt to Income Ratio vs Borrower Rte

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = DebtToIncomeRatio,
           y = BorrowerRate)) +
  geom_point(alpha = 1/25) +
  scale_x_continuous(
    lim = c(0, quantile(ld.no_na$DebtToIncomeRatio, 0.99)))
```

Observations: The correlation of Debt To Income Ratio and Borrower rate is weak
because we dont see any close to linear pattern.

We can conclude that Debt to Income Ratio has a interquartile range of 17%, and 
there is a sorprising max of 10.01. However since the mean is not very affected
by this outliers we know that the most of the values are relatively close to 
0.22

Also, since Prosper Rating has the greatest correlation with Borrower Rate, I 
wonder, how Debt to Income Ratio and the rating are related?


### 99 Percentile of Debt to Income Ratio vs ProsperRating..numeric.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Debt to income ratio is multiplied by 100 to obtain percentages and improve 
#quality of plot

ggplot(data = ld.no_na,
       aes(x = ProsperRating..factor., y = DebtToIncomeRatio* 100)) + 
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0,100,5), 
                     lim = c(0, quantile(ld.no_na$DebtToIncomeRatio* 100, 0.99)))
```

Observations: as the rating increases the debt to income ratio decreases, 
however, the difference between the rating 1 group mean and the rating 7 mean is 
only 10% greater, which is surprising because one might think that such 
different costumers will have a greater difference on this variable.

Continuing the analysis, I wanted how the debt variable changed over the years.


### 99 percentile of Debt to Income Ratio vs LoanOrigination
Date.year

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = LoanOriginationDate.year, y = DebtToIncomeRatio)) + 
  geom_boxplot() + 
  scale_y_continuous(breaks = seq(0,1,0.05), 
                     lim = c(0, quantile(ld.no_na$DebtToIncomeRatio, 0.99)))
```

Observations: we can see that the Debt to Income Ratio has stayed fairly similar
through the years, with a little increment the last years.

To continue I analyzed the Loan Original amount, I tought that with higher 
amounts the bank probably request more interest.


### Loan Original Amount with alpha = 1/10

```{r echo=FALSE, message=FALSE, warning=FALSE}
#LoanOriginalAmount is multiplied by 100 to obtain percentages and improve 
#quality of plot
ggplot(data = ld.no_na,
       aes(x = LoanOriginalAmount / 1000,
           y = BorrowerRate)) +
  geom_point(alpha = 1/25) +
  scale_y_continuous(breaks = seq(0,0.35,0.025)) +
  scale_x_continuous(breaks = seq(0,35,5))
```

Observations: my initial thoughts were incorrect, since the lower amounts have
greater intereset.


### Loan Origination Date vs Loan Origination Amount

```{r echo=FALSE, message=FALSE, warning=FALSE}
#LoanOriginalAmount is multiplied by 100 to obtain percentages and improve 
#quality of plot
ggplot(data = ld.no_na,
       aes(x = LoanOriginationDate.year, y = LoanOriginalAmount/1000)) + 
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0,36,2))
```

Observations: We can see that the amount have increased over time, we should 
remember from previous plot that in 2013 is where the majority of the loans were
originated, and now we know that also in the same year the bank loaned the 
greatest amounts.

Following the data set analysis, we analyze the bank loans' Estimated Return 

### Estimated return vs Borrower Rate with alpha of 1/10

```{r echo=FALSE, message=FALSE, warning=FALSE}
#EstimatedReturn is multiplied by 100 to obtain percentages and improve 
#quality of plot
ggplot(data = ld.no_na,
       aes(x = EstimatedReturn* 100,
           y = BorrowerRate)) +
  geom_point(alpha = 1/10) +
  scale_x_continuous(breaks = seq(-20,30,5))
```

Observations: as a whole, there is no strong correlation, however there are some
interesting lines, maybe another variable will help create linear relationships

Nonetheless, we should examine if this returns varies over the principal factor
of influence, the Prosper Rating:


### Prosper Rating factor vs Estimated Return boxplot

```{r echo=FALSE, message=FALSE, warning=FALSE}
#EstimatedReturn is multiplied by 100 to obtain percentages and improve 
#quality of plot
ggplot(data = ld.no_na,
       aes(x = ProsperRating..factor., y = EstimatedReturn * 100)) + 
  geom_boxplot() +
  scale_y_continuous(breaks = seq(-20,30,2.5))
```

Observations: the rating does influence the estimated return, we know 
that better ranking has lower BorrowerRates, and that is in accordance because
the bank is expecting lower returns for those ratings.

Continuing to the Employment Status Variable


### Employment Status vs Borrower Rate Boxplot

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = EmploymentStatus,
           y = BorrowerRate)) +
  geom_boxplot() + 
  scale_y_continuous(breaks = seq(0,0.35,0.025))
```

Observations: there is no clear relationship between the variables, the spreads
are mostly big, excluding not employed, that might be because the bank does not
issued for most of the unemployed costumers. We could check this with a bar 
graph of Employment Status Count.

To last variable of the bivariate analysis in the Income range. First, we look 
at the distribution for each category


### Income Range vs Borrower Rate Boxplots

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = IncomeRange,
           y = BorrowerRate)) +
  geom_boxplot() + 
  scale_y_continuous(breaks = seq(0,0.35,0.025))
```

Observations: as costumers earn more they get lower rates. However we should 
analyze how is the distributions of loans for each Income Rage group. Also we 
know not employed people don't usually get a loan, and the majority of the loans
are for people who earn 25000 - 75000. However there are over 25000 loans made 
to people who earn 75000 or more. We know that the majority of the loans are 
under 15000, and for that reason we can conclude that even people with high 
income request low amount loans.

Because of the decreasing Borrower Rate pattern of Income Ranges I wanted to see
if Income Range plays an important role on Prosper Rating.


### Income Range vs Prosper Rating Numeric

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = IncomeRange,
           y = ProsperRating..numeric.)) +
  geom_boxplot() + 
  scale_y_continuous(breaks = seq(1,7,1))
```

Observations: There are people with all the rating accros income range, even
thus the rating mean increases as the income does, there is no clear correlation.


# Bivariate Analysis

The result of the previous plots clearly shows that almost all of the variables 
have little to no effect on the borrower rate, this includes de date, the amount,
the employment status, and debt to income.

The principal factor is the prosper rating. From the client side, besides it 
only income range, loan amount and credit score seems to have some influence on 
the rate, decreasing it as the conditions are more favorable. 

In addition, the expected return plot is interesting, there seems to be some 
linear relationships on it, another variable might help clear this out.


# Multivariate Plots Section

Since it is clear that the main factor on the borrower rate is the prosper
rating, I will try to discover other relationships between this rating and the
other variables differencing each rating in the previous plots.


### Credit Score Average vs Borrower Rate by Prosper Rating with alpha = 1/10

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(data = ld.no_na,
       aes(x = CreditScoreAverage ,
           y = BorrowerRate)) +
  geom_point(position = 'jitter', alpha = 1/10, 
             aes(color = ld.no_na$ProsperRating..factor.)) + 
  scale_x_continuous(breaks = seq(610, 890, 20)) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Rating', reverse = T,
    override.aes = list(alpha = 1, size = 2)), palette = "Spectral")
```

Observations: As the credit score increases, so does the ranking, however the 
spreads are big.


### Loan Origination Date vs Borrower Rate by Prosper Rating alpha = 1/10

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = LoanOriginationDate.year,
           y = BorrowerRate)) +
  geom_point(alpha = 1/10, position = 'jitter' ,
             aes(color = ld.no_na$ProsperRating..factor.)) +
  scale_x_discrete(breaks = seq(2009,2014,1)) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Rating', reverse = T,
    override.aes = list(alpha = 1, size = 2)), palette = "Spectral")
```

Observations: we see that over the years the groups of ratings have kept a 
fairly similar borrower rate


### Debt to Income Ratio vs Borrower Rate by Prosper Rating

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = DebtToIncomeRatio,
           y = BorrowerRate)) +
  geom_point(alpha = 1/25, aes(color = ld.no_na$ProsperRating..factor.)) +
  scale_x_continuous(
    lim = c(0, quantile(ld.no_na$DebtToIncomeRatio, 0.99))) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Rating', reverse = T,
    override.aes = list(alpha = 1, size = 2)), palette = "Spectral")
```

Observations: Now that I included prosper rating as a factor, we know that even 
if the Debt to Income Ratio increases significantly, the borrower rate does not
change as much within groups.


### Loan Original Amount vs Borrower Rate by Prosper Rating with alpha = 1/10

```{r echo=FALSE, message=FALSE, warning=FALSE}
#LoanOriginalAmount is multiplied by 100 to obtain percentages and improve 
#quality of plot
ggplot(data = ld.no_na,
       aes(x = LoanOriginalAmount/1000 ,
           y = BorrowerRate)) +
  geom_point(aes(color = ProsperRating..factor.),
             alpha = 1/10, size = 1) + 
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Rating', reverse = T,
    override.aes = list(alpha = 1, size = 2)), palette = "Spectral") +
  scale_x_continuous(breaks = seq(0,40,2))
```

Observations: In this plot, we can corroborate that only higher rates get high 
loan amounts, however, costumers for each rating group request low amount loans.
One of the possible reasons behind this is that people with high income want
to improve their credit having a well payed low amount loan.


### Estimated Return vs Borrower Rate by Prosper Rating

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na, 
       aes(x = EstimatedReturn, y = BorrowerRate)) + 
  geom_point(aes(color = ProsperRating..factor.), alpha = 1/10) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Rating', reverse = T,
    override.aes = list(alpha = 1, size = 2)), palette = "Spectral")
```

Observations: this is a very interesting graphs of the analysis, because we can
see that for each Prosper Rating group, there is a linear relationship between
how much interest they pay and how much does the bank earns. This indicates that
the banks rating is indeed the most influential variable on rate, however for
the same group of people prosper set higher rates to obtain higher returns. It 
would be interesting to study this groups isolated to determine what makes the 
bank want to earn more or less within very similar financial profiles. In 
addition it is important to note that for the group of rating 1 there seems to 
be ar least 4 linear relationships, probably there is another variable that
is differentiating what the bank expects to return from them.


### Prosper Rating vs Borrower Rate by Income Range Boxplots

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = ProsperRating..factor.,
           y = BorrowerRate)) + geom_boxplot() +
  facet_wrap( ~ IncomeRange)
```

Observations: we see that across the Income Ranges the patter of negative 
correlations between rating and rate is really similar. 


### Prosper Rating vs Borrower Rate by Employment Status Boxplots

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ld.no_na,
       aes(x = ProsperRating..factor.,
           y = BorrowerRate)) + geom_boxplot() +
  facet_wrap( ~ EmploymentStatus)
```

Observations: As in the Income Range Boxplots, we see similar boxplots for each
Employment Status, however we know that the vast majority of the loans are for
employed costumers and for that reason the other boxplots have few data.


# Multivariate Analysis

First, we can conclude that neither LoanOriginationDate, DebtToIncomeRatio, 
LoanOriginalAmount, IncomeRange nor EmploymentStatus have any clear relationship
with the Borrower Rate.

Also, we can see that prosper rating is influenced by credit score average,
however, the spread is big, with very different credit scores having the same
rating.

In addition, we can see that over the years the rates for the ratings groups
was maintained.

Lower ratings tend to have bigger debt to income rating, however this is not a 
rule, there are people with rating 1 and close to 0 debt to income ratio.

Lower ratings tend to get lower loan amounts, and as the loan amount increases, 
we see that the amounts tend to be round numbers, there are several loans made 
with amount of 10,000, 15,000, 20,000, 25,000, 30,000 and 35,000.

There are linear relationships for the estimated return and borrower rate when
we group the data by rating, this describes that the bank calculates the
higher rate the higher the return will be, a relationship closes to +1. However,
we can also see that even with higher rates, the lower ratings are expected to
return very low. Especially costumers with prosper rating of 1 tend to create 
loses. In addition, there seems to be another variable that divides that 
group, because multiple linear relationships within the group.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(data = ld.no_na,
       aes(x = BorrowerRate)) + 
  geom_histogram(binwidth = 0.01, aes(fill = ProsperRating..factor.)) +
  scale_fill_brewer(type = 'div',
    guide = guide_legend(title = 'Prosper Rating', reverse = T,
    override.aes = list(alpha = 1, size = 2)), palette = "Spectral") +
  scale_x_continuous(breaks = seq(0,0.35,0.05))
```

### Description One

In the first Plot, I observe that the distribution of the borrower rates 
(interest) for the loans its similar to a normal distribution if rating 1 is
omitted, because almost of the costumer with this range get about 0.32 rate. 
Also, the other groups seem are in order and clearly separated, best ratings 
obtains bests rates. This is measured by the correlation, which is -0.953, a very
strong relationship


### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(data = ld.no_na,
       aes(x = CreditScoreAverage ,
           y = BorrowerRate)) +
  geom_point(position = 'jitter', alpha = 1/25, 
             aes(color = ld.no_na$ProsperRating..factor.)) + 
  scale_x_continuous(breaks = seq(610, 890, 20)) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Prosper Rating', reverse = T,
    override.aes = list(alpha = 1, size = 2)), palette = "Spectral") +
  scale_y_continuous(breaks = seq(0,0.35,0.025))
```

### Description Two

This plot is interesting because it demonstrated that better credit scores tend 
to get better borrower rates, however the relationship is not strong with a 
correlation of -0.529. Also, the fact that same credit score averages tend to 
have several costumers with different ratings. For example, we can find all the
ratings in good amount in the 730-credit score average bin. This is very
surprising for me because I thought that this will be a very influential 
variable on borrower rate and prosper rating. We can conclude that there are 
other variables influencing the rating and borrower rate.


### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(data = ld.no_na, 
       aes(x = EstimatedReturn, y = BorrowerRate)) + 
  geom_point(aes(color = ProsperRating..factor.), 
             alpha = .33) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Prosper Rating', reverse = T,
    override.aes = list(alpha = 1, size = 2)), palette = "Spectral") +
  scale_y_continuous(breaks = seq(0,0.35,0.05)) + 
  scale_x_continuous(breaks = seq(-0.2,0.25,0.05))
```

### Description Three

This final plot stands out because even thus I did not find any linear 
relationships in the other variables, here we see that if we divide each group
of ratings, we could create several linear relationships. 

This describes that the bank calculates that the higher rate the return will be 
higher too, a relationship close to +1. However, we can also see that this is 
the case for the ratings besides 1, where even the higher ratings are expected 
to return lower. 

Also, the bank seems to set very different rates for the people with similar
financial characteristics, there must be another variable that influence how
much prosper wants to earn. So if someone wants to obtain a lower rate, even
thus rating is the most influential variable, is not the only one they should 
take into consideration.

Finally costumers with prosper rating 1 tend to create loses. Also, there 
there seems to be another variable that divides that group, because multiple
linear relationships may be within the group.

------


# Reflection

Data analysis is a process where you first you have to understand each variable 
and the problem before doing any plots, because if you don't understand the 
situation you can't formulate the questions that will get the answers you need. 

It is very important to clean up the data, because some variables might have 
classes or formats that will produce bad visualizations and findings. After that
decide which is your main characteristic of interest, what are you trying to 
predict and study that variable alone, only after you have an understanding of 
it you can explore the relationships it has with the other variables. 

Create plots for all the variables and try to find key findings that will lead 
your investigation until you find an explanation of the phenomenon’s discovered. 

On a final note, I believe that the data is as valuable as it is real or 
trustworthy. Even thus is it from a great source the analyst must be aware of the
limitations of the data due to how was is recollected, from who and under what 
circumstances.

Regarding this data set, it would be valuable to study the rest of the variables
to discover what influences Porsper rating, because if it can be predicted we 
could predict the loan characteristics better. Also, it would be interesting to 
compare the estimated return with the actual return of the loans.